---
title: "Introduction to HEART"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to HEART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

HEART tests the association between each variant and a phenotype **in a target
population**. It borrows information from a larger source population whose
genetic effects may differ. This vignette walks through the two steps of the
method on simulated data.

```{r}
library(HEART)
sim <- heart_simulate(n_tar = 3000, n_src = 20000, J = 3000, seed = 2026)
table(sim$truth$config)
```

The configurations are `00` (null in both populations), `01` (source-specific
signal, so the target null holds), `10` (target-specific signal) and `11` (shared
signal).

## Step 1: target-only, auxiliary and decorrelated estimators

```{r}
fit <- heart_assoc(sim$Y, sim$G, sim$D, sim$X, aux = c("pooled", "source"))
head(fit[, c("snp", "beta_tar", "p_tar", "beta_pooled", "rho_pooled",
             "beta_de_pooled", "p_de_pooled")])
```

The pooled estimator is not valid for the target null. Its p-values are small
for the source-specific variants (`01`), whose target effect is zero:

```{r}
src_only <- sim$truth$config == "01"
c(pooled = mean(fit$p_pooled[src_only] < 1e-4),
  target = mean(fit$p_tar[src_only] < 1e-4))
```

## Step 2: HEART test

```{r}
res <- heart_test(fit$p_tar, fit$p_de_pooled, snp = fit$snp, alpha = 0.05)
res
```

Compare the discoveries with the truth:

```{r}
signal <- sim$truth$beta_tar != 0
bh <- p.adjust(fit$p_tar, "BH") <= 0.05
rbind(
  HEART = c(discoveries = sum(res$results$reject_fdr),
            FDP = mean(!signal[res$results$reject_fdr]),
            power = mean(res$results$reject_fdr[signal])),
  `Target-only BH` = c(sum(bh), mean(!signal[bh]), mean(bh[signal]))
)
```

## Inspecting the null distribution

`heart_null_cdf()` returns the components of the estimated composite null
distribution: F<sub>00</sub>, Q̂, F̂<sub>01</sub>, N̂ and F̂<sub>0</sub>.

```{r}
nc <- res$null_cdf
head(nc)
```
